Using Random Walk Simulations to Calculate Ground State Energies in Quantum Physics

### 2020-04-19 Paper/main.tex
Completed Results and Conclusion sections

\begin{figure}%[H]%[!ht]
  \begin{center}
    \includegraphics[width=0.45\textwidth]{images/cartoon.pdf}
    \caption{Illustrations, graphs, and photographs may fit across both columns, if necessary. Your artwork must be in place in the article.}
    \label{fig:cartoon}
  \end{center}
\end{figure}

@@

The random walk ends when a walker dies, with a
specifically chosen $a(j)$. Initially, the $a(j)$ is chosen to be that of an
infinite square well \todo{cite}, that is:
\begin{equation}
  \label{eq:squarewell}
    a(j) =
    \begin{cases}
      0,& \text{if } j < J\\
      \infty,& \text{if } j \geq J
    \end{cases}
    \nonumber
\end{equation}
where J is a well-defined boundary. So the walker has no chance of dying whilst
within the well, but will die immediately once reaching the boundary. As
explained in Section \ref{sec:TheoreticalBackground}, we expect that given
sufficient time, all the walkers will reach the boundary and die. But in order
to save computational resources, a maximum number of steps may also be defined,
such that a walker will also die should they move that number of steps without
reaching a boundary. This was not initially implemented, as it can introduce
systemic errors, since there would be a disproportionate number of walkers dying
on the maximum number of steps.

@@

To find the rate of death $\lambda$, we consider Equation \ref{eq:firsteq}. With
some simple manipulation:
\begin{equation}
  \begin{split}
    p(j, n) & = q(j) \exp(-\lambda n)\\
    \log(p) & = \log(q \exp(-\lambda n)) \\
    \log(p) & = -\lambda n + log(q)
  \end{split}
  \nonumber
\end{equation}
Thus, by plotting the logarithm of the probability of surviving $n$ steps
against the number of steps $n$, we can obtain the death rate $\lambda$ from the
gradient.

@@

\section{Results and Discussion}

\begin{figure*}[ht!]
  \centering
  \subfloat[Histogram of probability of dying on particlar step]
  {\includegraphics[width=0.5\textwidth]{images/exp_plot.pdf}
  \label{fig:hist}}
  \centering
  \subfloat[Histogram of probability of dying on particular step in a logarithmic scale]
  {\includegraphics[width=0.5\textwidth]{images/line_plot.pdf}
    \label{fig:loghist}}
  \caption{Histograms showing the Probability of dying on a particular step in
    an infinite square well with $J = 20$. As can be seen, the peak is away from the
    origin, as is expected since it is not likely for a walker to die on a step $n$
    where $n < J$. The exponential decay after the peak is evident, implying a
    straight line shape if plotted on a logarithmic scale, which can be seen on
    the right hand side plot. In this logarithmic scale plot, the peak being away
    from the origin is clearer.}
  \label{fig:expplots}
\end{figure*}

\begin{figure*}[ht!]
  \centering
  \subfloat[Histogram of probability of surviving to at least a particlar step]
  {\includegraphics[width=0.5\textwidth]{images/cum_exp_plot.pdf}
  \label{fig:cumhist}}
  \centering
  \subfloat[Histogram of probability of surviving to at least a particular step
  in a logarithmic scale]
  {\includegraphics[width=0.5\textwidth]{images/cum_line_plot.pdf}
  \label{fig:cumloghist}}
  \caption{Histogram showing the Probability of surviving to at least a
      particular step in an infinite square well with $J = 20$. This is a
      cumulative sum of the bins in Figure \ref{fig:hist}. As can be seen, the
      peak is at the origin, since it is certain that no walker will die on the
      first step. An exponential decay curve provides a good fit, with an $R^2$
      value of 0.9999998, very close to 1. Plotting on a logarithmic scale shows
      the straight line fit, the gradient of which is $\lambda$, giving $\lambda
      J^2 = 1.2334$, with a negligible error.}
  \label{fig:cumplots}
\end{figure*}

\begin{figure}[!ht]
  \begin{center}
    \includegraphics[width=0.45\textwidth]{images/multiplot.pdf}
    \caption{Histogram plotted on a logarithmic scale showing the Probability of
      surviving to at least a particualr step in different infinite square wells with
      different boundaries $J$. The $\lambda J^2$ values for each are shown. The
      values get closer to the theoretical value of $\frac{\pi^2}{8}$ as we increase
      the size of the well. This is likely due to more walkers surviving to a higher
      number of steps, as the well size is increased.}
    \label{fig:multi_line_plot}
  \end{center}
\end{figure}

\begin{figure}[!ht]
  \begin{center}
    \includegraphics[width=0.45\textwidth]{images/cum_line_plot_cutoff.pdf}
    \caption{Histogram plotted on a logarithmic scale showing the Probability of
      surviving to at least a particular step in an infinite square well with $J
      = 20$, and also introducing a maximum number of steps at 2000, after which
      all walkers will die. This saves computational resources but introduces
      some systematic error, as there will now be a disproportionately high
      number of walkers in the 2000 steps bin.}
    \label{fig:cum_line_cutoff_plot}
  \end{center}
\end{figure}

@@

Sample Text

>>>>

\begin{figure}%[H]%%[!ht]
  \begin{center}
    \includegraphics[width=0.45\textwidth]{images/cartoon.pdf}
    \caption{Illustrations, graphs, and photographs may fit across both columns,
      if necessary. Your artwork must be in place in the article.}
    \label{fig:cartoon}
  \end{center}
\end{figure}

@@

The random walk ends when a walker dies, with a
specifically chosen $a(j)$. Initially, the $a(j)$ is chosen to be that of an
infinite square well \todo{cite}, that is:
\begin{equation}
  \label{eq:squarewell}
    a(j) =
    \begin{cases}
      0,& \text{if } j < J\\
      \infty,& \text{if } j \geq J
    \end{cases}
    \nonumber
\end{equation}
where J is a well-defined boundary. So the walker has no chance of dying whilst
within the well, but will die immediately once reaching the boundary. As
explained in Section \ref{sec:TheoreticalBackground}, we expect that given
sufficient time, all the walkers will reach the boundary and die. But in order
to save computational resources, a maximum number of steps may also be defined,
such that a walker will also die should they move that number of steps without
reaching a boundary. This was not initially implemented, as it can introduce
systematic errors, since there would be a disproportionate number of walkers dying
on the maximum number of steps.

@@

To find the rate of death $\lambda$, we consider Equation \ref{eq:firsteq}. With
some simple manipulation:
\begin{equation}
  \begin{split}
    p(j, n) & = q(j) \exp(-\lambda n)\\
    \log(p) & = \log(q \exp(-\lambda n)) \\
  \end{split}
  \nonumber
\end{equation}

\begin{equation}
  \log(p) = -\lambda n + log(q)
  \label{eq:straightline}
\end{equation}

Thus, by plotting the logarithm of the probability of surviving $n$ steps
against the number of steps $n$, we can obtain the death rate $\lambda$ from the
gradient.

@@

\section{Results and Discussion}
The simulation was first run for an infinite square well with boundary $J = 20$,
with 1000000 walkers. The step on which each walker died was recorded and then
the normalised histogram in Figure \ref{fig:hist} was constructed. The histogram
has been normalised, and is hence a probability distribution showing the
probability that a walker will die on a particular step. As we expect, the peak
is far from the origin, since for a walker to die in an inifinite square well
with $J=20$, they would have to walk at least 20 steps. It is also highly
unlikely that a walker would walk 20 steps in the same direction
consecutively. Thus, the peak is also at a number of steps greater than $J$. We
expect the shape of the decay after the peak to be exponential. We can see this
by plotting the histogram on a logarithmic scale.

Figure \ref{fig:loghist} presents the same histogram data with a logarithmic y
scale. In this figure, a straight line is evident in the data, showing the decay
was indeed exponential after the peak. The peak probability being away fro the
origin is also clearer in this figure. Figure \ref{fig:loghist} shows a
``widening'' of the straight line at higher step numbers. This is because there
is a greater variance in the number of walkers surviving to this number as it is
highly unlikely.

By taking the cumulative sum of the histograms in Figure \ref{fig:expplots}, we
can ensure the peak probability begins at the origin. Figure \ref{fig:cumplots}
present the cumulative sums of the histograms. They represent the probability of
surviving at least a particular number of steps. Hence, the peak probability of
1 is at the origin, since there is a probability of 1 that a walker will survive
the first step.

Figure \ref{fig:cumhist} presents the cumulative summed histogram, with a fitted
exponential curve of the form of Equation \ref{eq:firsteq}. The fit is very
good, with $R^2 = 0.9999998$, very close to 1. There is also negligible
numerical error in the fit, i.e. an error smaller than the quoted precision of
the fit parameters. This is because the simulation was run for 1000000 walkers,
providing a large sample size. Figure \ref{fig:cumloghist} presents the same
histogram with a logarithmic y-scale and a fitted straight line of the form of
Equation \ref{eq:straightline}. This fit has the same $R^2$ value, as it is
simply the same fit on a different y-scale, and is thus equally good.

We observe some fluctuation in the data in the higher steps. This is because
there are so few walkers surviving to this many steps that there is more
significant variance in their relative numbers. It is possible to remove this
variance by introducing a cutoff number of steps, i.e. a maximum number of steps
at which the walker will automatically die, regardless of whether they reached a
boundary or not. Thus, there will not be such high variance in the number of
walkers surviving to high number of steps. This is presented in Figure
\ref{fig:cum_line_cutoff_plot}. While this fit is observed to be better, with a
higher $R^2$ value, some systematic error has been introduced since there will now
be a disproportionately high number of walkers dying at the cutoff step. Thus,
the final $\lambda J^2$ value will be less accurate. On a faster computer, with
more efficient code, it is possible to forgo the step cutoff and allow the
simulation to run until all the walkers die. This may take much longer but will
ensure there will not be such a systematic error introduced.

We can use the gradient of the fitted line in Figure \ref{fig:cumloghist} as the
$\lambda$ value, and compare $\lambda J^2$ with the theoretical value. However,
to determine the reliability of the calculated value, it is necessary to repeat
the procedure many times and take the average $\lambda$ and the standard
deviation as the error.

It is also necessary to investigate how varying the boundary $J$ affects the
accuracy of the final answer. Figure \ref{fig:multi_line_plot} presents the
cumulative summed histograms with their fitted lines for many different boundary
$J$ values. No step cutoff was introduced. The calculated $\lambda J^2$ value is
shown in the legend. \todo{truncate numbers} As shown, the gradient changes, but
this is to be expected since the expected highest number of steps would increase
as the boundary is larger, since there is more ``space'' for the walker to move
before reaching a boundary. As shown in the legend, the calculated $\lambda J^2$
values get closer and closer and closer to the theoretical value as $J$
increases. This is due to the fact that as $J$ increases, there are more walkers
surviving to larger number of steps. \todo{make plot of lambda J against J}.
Thus, we can say that using a larger value of $J$ will result in a more
accurate answer but will also be more computationally intensive and take
longer. Thus, we make a tradeoff between speed and accuracy and choose $J=20$
for the final investigation.

Running the simulation 10 times and taking the average $\lambda J^2$ with error
being standard deviation, we find: $\lambda J^2 = 1.2339 \pm 0.0008$. We quote
the precision to the decimal place at which there is variation from the
theoretical result of $1.2337$. We notice also that the error in the calculated
answer is small (since there is not much variation between repeated simulations)
and the theoretical result is within the range of the calculated answer.

\todo{circular well}
\begin{figure*}%[ht!]
  \centering
  \subfloat[Histogram of probability of dying on particlar step]
  {\includegraphics[width=0.5\textwidth]{images/exp_plot.pdf}
  \label{fig:hist}}
  \centering
  \subfloat[Histogram of probability of dying on particular step in a logarithmic scale]
  {\includegraphics[width=0.5\textwidth]{images/line_plot.pdf}
    \label{fig:loghist}}
  \caption{Histograms showing the Probability of dying on a particular step in
    an infinite square well with $J = 20$. As can be seen, the peak is away from the
    origin, as is expected since it is not likely for a walker to die on a step $n$
    where $n < J$. The exponential decay after the peak is evident, implying a
    straight line shape if plotted on a logarithmic scale, which can be seen on
    the right hand side plot. In this logarithmic scale plot, the peak being away
    from the origin is clearer.}
  \label{fig:expplots}
\end{figure*}

\begin{figure*}%[ht!]
  \centering
  \subfloat[Histogram of probability of surviving to at least a particlar step]
  {\includegraphics[width=0.5\textwidth]{images/cum_exp_plot.pdf}
  \label{fig:cumhist}}
  \centering
  \subfloat[Histogram of probability of surviving to at least a particular step
  in a logarithmic scale]
  {\includegraphics[width=0.5\textwidth]{images/cum_line_plot.pdf}
  \label{fig:cumloghist}}
  \caption{Histogram showing the Probability of surviving to at least a
      particular step in an infinite square well with $J = 20$. This is a
      cumulative sum of the bins in Figure \ref{fig:hist}. As can be seen, the
      peak is at the origin, since it is certain that no walker will die on the
      first step. An exponential decay curve provides a good fit, with an $R^2$
      value of 0.9999998, very close to 1. Plotting on a logarithmic scale shows
      the straight line fit, the gradient of which is $\lambda$, giving $\lambda
\todo{fix centering of cum plot}
      J^2 = 1.2334$, with a negligible error.}
  \label{fig:cumplots}
\end{figure*}

\begin{figure}%[!ht]
  \begin{center}
    \includegraphics[width=0.45\textwidth]{images/cum_line_plot_cutoff.pdf}
    \caption{Histogram plotted on a logarithmic scale showing the Probability of
      surviving to at least a particular step in an infinite square well with $J
      = 20$, and also introducing a maximum number of steps at 2000, after which
      all walkers will die. This saves computational resources but introduces
      some systematic error, as there will now be a disproportionately high
      number of walkers in the 2000 steps bin.}
    \label{fig:cum_line_cutoff_plot}
  \end{center}
\end{figure}

\begin{figure}%[!ht]
  \begin{center}
    \includegraphics[width=0.45\textwidth]{images/multiplot.pdf}
    \caption{Histogram plotted on a logarithmic scale showing the Probability of
      surviving to at least a particualr step in different infinite square wells with
      different boundaries $J$. The $\lambda J^2$ values for each are shown. The
      values get closer to the theoretical value of $\frac{\pi^2}{8}$ as we increase
      the size of the well. This is likely due to more walkers surviving to a higher
      number of steps, as the well size is increased.}
    \label{fig:multi_line_plot}
  \end{center}
\end{figure}

@@

In this investigation we demonstrated a method of calculating ground state
energies of a quantum system using Random Walk computational simulations. The
investigation focussed on an Infinite Square Well, but is easily extended to a
Symmetric potential. The simualation was written in C++ to maximise performance
and is available at \todo{cite}.

The investigation explored the effects of varying the size of the well in
determining the ground state energy of a quantum system, and showed that this
method is more effective for larger wells, i.e. wells with larger
boundaries. This is because for wells with larger boundaries, the walker can
take a larger number of steps on average before dying, and thus can populate
larger step bins. Thus, this method produces data that is close to the
theoretical result for larger infinite square wells.

The general theoretical result for the ground state energy of an infinite square
well is $\frac{\pi^2}{8}\frac{\hbar^2}{mJ^2}$ where $J$ is the boundary of the
well, and this method allows for the calculation of $\lambda R^2
\frac{\hbar^2}{mJ^2}$, where $R$ also corresponds to the boundary of the
well. Thus, we expect our calculated result to be $\lambda R^2 = \lambda J^2 =
\frac{\pi^2}{8}$.

This method provided $\lambda J^2 = 1.2339 \pm 0.0008$, which is very close to
the theoretical result.

This method can be easily extended to multiple dimensions with only a linear
increase in time complexity for each added dimension. Thus, while this method
may not be very efficient for low number of dimensions, it quickly becomes
effective at a higher number of dimensions.
